The locally Gaussian partial correlation

### 2020-09-09 lgpc_JBES.Rmd
Fixing new round of corrections from Dag

---
title: "The locally Gaussian partial correlation"
author: Håkon Otneim^[Corresponding author. Department of Business and Management Science, NHH Norwegian School of Economics. Helleveien 30, 5045 Bergen, Norway. \texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen]
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: yes
      toc: no
      includes:
          in_header: latex_preamble.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
abstract: > 
   It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally,  we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
---

@@

Definitions and properties are presented in Section \@ref(chap:conditional), whereas estimation is covered in Section \@ref(chap:estimation). Two examples illustrating the strengths of our approach are given in Section \@ref(chap:example). Section \@ref(chap:testing) gives a brief glimpse of the potential in testing.

An integral part of our article is the online supplement. It deals with a number of issues complementing the treatment in the main paper. Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of Pitman efficiency for testing and validity of the use of the bootstrap.

@@

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
(\#eq:trans)
\end{equation}
The transformation enables us to simplify the local Gaussian approximation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$  at the point $\fv = \z$ as 
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
(\#eq:lgdeapprox)
\end{equation}  
where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a further simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.

In practice, we do not know $F_{j}$, but we can instead use the empirical distribution function 
$$F_{j,n}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),$$
where $1(\cdot)$ is the indicator function, $n$ is the number of observations, and $X_{ji}$ is the $i$th observation of $X_j$. This results in pseudo standard normal variables $Z_{j,n} = \Phi^{-1}(F_{j,n}(X_j))$.

In the following, we will not always distinguish between $Z_j$ and $Z_{j,n}$. In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $Z_{j,n}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.

We will in this paper refer to $\X$ and its probability density function $f_{\X}$ as being on the \(x\)-scale, and to $\Z$ and its probability density function $f_{\Z}$ as being on the \(z\)-scale. For further discussion of the simplified $z$-representation we refer to @otneim2017locally.

@@

It is of course possible to introduce an LGPC $\alpha(\x)$ directly on the $x$-scale, but that representation is in many ways harder to handle both computationally and asymptotically. For a multivariate Gaussian distribution, we have that $\alpha_{\X}(\x) = \alpha_{\Z}(\z) = \alpha$. In the remainder of this paper, we write mainly in terms of the $z$-representation using the LGPC $\alpha(\z) = \alpha_{\Z}(\z)$, but when we write the local partial correlation between $X_1$ and $X_2$ given $\X_3 = \x_3$ at the point $(x_1,x_2,\x_3)$, this is simply $\alpha(\z)$ inserted with $z_j = \Phi^{-1}(F_{j}(x_j))$, $j=1,\ldots,p$.

@@

Then, by writing $b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0$ assuming they all converge to zero at the same rate, 
\begin{equation}
\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
(\#eq:an3)
\end{equation}
where $\bm{I}$ is the $3\times3$ identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
```

@@

Then, by letting $b_n^2$ (assuming identical convergence rates) mean the product $b_{nj}b_{nk}$ corresponding to the pairs of variables defining the components in $\frho$, -->
\begin{equation}
\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fOmega),
\label{eq:ot2018}
\end{equation}
<!--
where $\fOmega$ is a diagonal matrix with components
\begin{equation}
\fOmega^{(\ell,\ell)} = 
\frac{f_{\ell}(\z_{\ell})\int K^2(\y_{\ell})\,\textrm{d}\y_{\ell}}
{u^2(\z_{\ell},\rho_{0,\ell}(\z_{\ell}))}
\psi^2(\z_{\ell}, \rho_{0,\ell}(\z_{\ell})),
(\#eq:asymcov)
\end{equation}
where $\ell = 1,\ldots,p(p-1/2)$ runs over all pairs of variables, and $f_{\ell}(\cdot)$ is the bivariate marginal density function of the pair $\Z_{\ell}$.
```

@@

For a proof, we refer to @otneim2017conditional.--> where the asymptotic covariance matrix $\fOmega$ is defined in an analogous way as in Theorem \@ref(thm:loccor3) There are two important differences between these two results, however. First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of $1/(nb^2)^{1/2}$. This is because we estimate the local correlations one by one based on pairs of variables. The second difference is that the leading term of the asymptotic covariance matrix in Theorem 2 by @otneim2017conditional has a simple analytic expression that can be used to construct, for instance, confidence intervals. We refer to Section A.3 in the online supplement for further details regarding the limiting distribution of the LGPC as well conditions for consistency of the estimates.

@@

\begin{equation}
\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
(\#eq:limitlgpc)
\end{equation}
where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fOmega$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate). In any case, it is only a matter of basic differentiation to work out an expression for $\nabla g(\cdot)$, and this task has been deferred to Section C of the online supplement in the case where $d_1 = d_2 = 1$.

@@

\begin{equation}
R_{1,t} \perp R_{2, t-1} \,\,|\,\, R_{1,t-1},
\end{equation}
which we may test formally using the approaches described in the following section or the references therein. Let us rather concentrate on *describing* this conditional dependence relationship by estimating the local Gaussian partial correlation between $R_{1,t}$ and $R_{2,t-1}$ given a few different values of $R_{2,t-1}$, which gives considerable more information than a single $p$-value of a test. We collect daily observations on Oslo Stock Exchange^[Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.] from 1 January 2018 until 31 December 2019. In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal $R_{1,t} = R_{2,t-1}$ for three different values $R_{1,t-1} = -0.02, 0$ and $0.02$, along with their block-bootstrapped 95% confidence bands. The constant $\alpha(\z) = 0$ is indicated as a dotted line.

@@

One could perhaps make the argument that $R_{1,t}$ and $R_{2,t-1}$ seem to be somewhat positively dependent in the right half of the graph for *small* values of the conditioning variable $R_{1,t-1}$. Roughly translated to this particular situation this means that if the large-firm portfolio gives a negative return on day $t-1$, then its value on day $t$ depend on the performance small-firm portfolio on day $t-1$ in a nonlinear and non-Gaussian way that is characterized by positive dependence in the first quadrant. The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence bands include zero in most of the sample space.

@@

We construct a test statistic for testing

$$\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3$$
or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,

\begin{equation}
\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3
(\#eq:ci-test)
\end{equation}
by aggregating our local measure of dependence over the sample space of $\X$ (or $\Z$). A natural test statistic on the $z$-scale is

@@

Departures from conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test). Asymptotically, one might expect that approximate $p$-values for our test can be extracted from the limiting distribution of the test statistic $T_{n,b}$, which we derive in the online supplement Section F. The limiting asymptotic distribution of $T_{n,b}-T$ is given in Theorem F.2 for the 1-parameter general pairwise case, whereas Theorem F.3 treats the 3-coordinate (3-parameter) case with $d_1=d_2=d_3=1$. It is interesting to note from the discussion immediately following the proofs of these theorems that the convergence rate in the first case is $n^{-1/2}$, and in the second case is $\left(nb^2\right)^{-1/2}$. This means that the tests are asymptotically able to pick up alternatives approaching the null hypothesis of conditional independence at these rates. This is measured by the Pitman criterion. A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of @wang2017characteristic. The rate $n^{-1/2}b^{-d_3/4}$ of Wang and Hong compares favorable with other tests. The rate $n^{-1/2}$ obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive. We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.

@@

There is a potential for much further work. A next natural step may be structural equations as used in network analysis and Pearl-type causality [@pearl2000causality], and ultimately further exploring the relationship or lack of relationship between that type of causality and Granger causality.

All methods presented in this paper have been implemented in the R programming language [@r], and is available in the package **lg** [@otneim2019lg]. In addition, we provide code and data for easy replication of the results presented here in an online appendix.

**Acknowledgements.** The authors gratefully acknowledge the input of an associate editor as well as two anonymuos referees, whose comments have greatly improved this paper.

>>>>

---
title: "The locally Gaussian partial correlation"
author: Håkon Otneim^[Corresponding author. Department of Business and Management Science, NHH Norwegian School of Economics. Helleveien 30, 5045 Bergen, Norway. \texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen \newline {Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.]
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: yes
      toc: no
      includes:
          in_header: latex_preamble.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
abstract: > 
   It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally,  we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
---

@@

Definitions and properties are presented in Section \@ref(chap:conditional), whereas estimation is covered in Section \@ref(chap:estimation). Two examples illustrating the strengths of our approach are given in Section \@ref(chap:example). Section \@ref(chap:testing) gives a brief account of the potential in testing.

An integral part of our article is the online supplement. It deals with a number of issues complementing the treatment in the main paper. Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of local asymptotic power and the Pitman criterion for testing and validity of the use of the bootstrap.

@@

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
(\#eq:trans)
\end{equation}
The transformation enables us to simplify the local Gaussian approximation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$  at the point $\fv = \z$ as 
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
(\#eq:lgdeapprox)
\end{equation}  
where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a further simplified approximation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.

In practice, we do not know $F_{j}$, but we can instead use the empirical distribution function 
$$\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),$$
where $1(\cdot)$ is the indicator function, $n$ is the number of observations, and $X_{ji}$ is the $i$th observation of $X_j$. This results in pseudo standard normal variables $\widehat Z_{j} = \Phi^{-1}(\widehat F_{j}(X_j))$.

In the following, we will not always distinguish between $Z_j$ and $\widehat Z_{j}$. In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $Z_{j,n}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.

We will in this paper refer to $\X$ and its probability density function $f_{\X}$ as being on the \(x\)-scale, and to $\Z$ and its probability density function $f_{\Z}$ as being on the \(z\)-scale. For further discussion of the simplified $z$-approximation we refer to @otneim2017locally.

@@

It is of course possible to introduce an LGPC $\alpha(\x)$ directly on the $x$-scale, but that representation is in many ways harder to handle both computationally and asymptotically. For a multivariate Gaussian distribution, we have that $\alpha_{\X}(\x) = \alpha_{\Z}(\z) = \alpha$. In the remainder of this paper, we write mainly in terms of the $z$-representation using the LGPC $\alpha(\z) = \alpha_{\Z}(\z)$, but when we write the local partial correlation between $X_1$ and $X_2$ given $\X_3 = \x_3$ at the point $(x_1,x_2,\x_3)$, this is simply $\alpha(\z)$ inserted $z_j = \Phi^{-1}(F_{j}(x_j))$, $j=1,\ldots,p$.

@@

Then, by writing $b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0$ assuming they all converge to zero at the same rate, 
\begin{equation}
\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
(\#eq:an3)
\end{equation}
where $\bm{I}$ is the $3\times3$ identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
```

@@

Then, by letting $b_n^2$ (assuming identical convergence rates) mean the product $b_{nj}b_{nk}$ corresponding to the pairs of variables defining the components in $\frho$, -->
\begin{equation}
\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
\label{eq:ot2018}
\end{equation}
<!--
where $\fOmega$ is a diagonal matrix with components
\begin{equation}
\fOmega^{(\ell,\ell)} = 
\frac{f_{\ell}(\z_{\ell})\int K^2(\y_{\ell})\,\textrm{d}\y_{\ell}}
{u^2(\z_{\ell},\rho_{0,\ell}(\z_{\ell}))}
\psi^2(\z_{\ell}, \rho_{0,\ell}(\z_{\ell})),
(\#eq:asymcov)
\end{equation}
where $\ell = 1,\ldots,p(p-1/2)$ runs over all pairs of variables, and $f_{\ell}(\cdot)$ is the bivariate marginal density function of the pair $\Z_{\ell}$.
```

@@

For a proof, we refer to @otneim2017conditional.--> where the asymptotic covariance matrix $\fLambda$ is defined as $\fSigma$ in Theorem 2 of @otneim2017conditional. There are two important differences compared to Theorem \@ref(thm:loccor3), however. First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of $1/(nb^2)^{1/2}$. This is because we estimate the local correlations one by one based on pairs of variables. The second difference is that the leading term of the asymptotic covariance matrix in Theorem 2 by @otneim2017conditional has a simple analytic expression that can be used to construct, for instance, confidence intervals. We refer to Section A.3 in the online supplement for further details regarding the limiting distribution of the LGPC as well conditions for consistency of the estimates.

@@

\begin{equation}
\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
(\#eq:limitlgpc)
\end{equation}
where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fLambda$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate). In any case, it is only a matter of basic differentiation to work out an expression for $\nabla g(\cdot)$, and this task has been deferred to Section C of the online supplement in the case where $d_1 = d_2 = 1$.

@@

\begin{equation}
R_{1,t} \perp R_{2, t-1} \,\,|\,\, R_{1,t-1},
\end{equation}
which we may test formally using the approaches described in the following section or the references therein. Let us rather concentrate on *describing* this conditional dependence relationship by estimating the local Gaussian partial correlation between $R_{1,t}$ and $R_{2,t-1}$ given a few different values of $R_{2,t-1}$, which gives considerable more information than a single $p$-value of a test. We collect daily observations on Oslo Stock Exchange^[Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.] from 1 January 2018 until 31 December 2019. In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal $R_{1,t} = R_{2,t-1}$ for three different values $R_{1,t-1} = -0.02, 0$ and $0.02$, along with their block-bootstrapped 95% confidence intervals. The constant $\alpha(\z) = 0$ is indicated as a dotted line.

@@

One could perhaps make the argument that $R_{1,t}$ and $R_{2,t-1}$ seem to be somewhat positively dependent in the right half of the graph for *small* values of the conditioning variable $R_{1,t-1}$. Roughly translated to this particular situation this means that if the large-firm portfolio gives a negative return on day $t-1$, then its value on day $t$ depend on the performance small-firm portfolio on day $t-1$ in a nonlinear and non-Gaussian way that is characterized by positive dependence in the first quadrant. The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence intervals include zero in most of the sample space.

@@

We construct a test statistic for testing $\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3,$ or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,

\begin{equation}
\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3,
(\#eq:ci-test)
\end{equation}
by aggregating our local measure of dependence over the sample space of $\X$ (or $\Z$). A natural test statistic on the $z$-scale is

@@

Departures from conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test). Asymptotically, one might expect that approximate $p$-values for our test can be extracted from the limiting distribution of the test statistic $T_{n,b}$, which we derive in the online supplement Section F. The limiting asymptotic distribution of $T_{n,b}-T$ is given in Theorem F.2 for the 1-parameter general pairwise case, whereas Theorem F.3 treats the 3-coordinate (3-parameter) case with $d_1=d_2=d_3=1$. It is interesting to note from the discussion immediately following the proofs of these theorems that the convergence rate in the first case is $n^{-1/2}$, and in the second case is $\left(nb^2\right)^{-1/2}$. This means that the tests are asymptotically able to pick up alternatives approaching the null hypothesis of conditional independence at these rates. This is measured by the Pitman criterion. A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of @wang2017characteristic. The rate $n^{-1/2}b^{-d_3/4}$ of Wang and Hong compares favorable with other tests. The rate $n^{-1/2}$ obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive. We go into more detail regarding local asymptotic power and the Pitman criterion in Section G of the online supplement.

@@

There is a potential for much further work. A next natural step may be structural equations as used in network analysis and Pearl-type causality [@pearl2000causality], and ultimately further exploring the relationship or lack of relationship between that type of causality and Granger causality. All methods presented in this paper have been implemented in the R programming language [@r], and is available in the package **lg** [@otneim2019lg]. In addition, we provide code and data for easy replication of the results presented here in an online appendix.
